---
title: K-means and Hierarchical Clustering for Exploratory Data Analysis
author: ''
date: '2019-12-23'
slug: k-means-and-hierarchical-clustering-for-exploratory-data-analysis
categories: []
tags:
  - machine learning
  - R
summary: ''
authors: []
external_link: ''
image:
  caption: ''
  focal_point: ''
  preview_only: no
url_code: ''
url_pdf: ''
url_slides: ''
url_video: ''
slides: ''
---


This analysis applies two common clustering techniques, k-means clustering and hierarchical clustering, to data on house prices in England and Wales. It is intended to demonstrate that fitting simple models to a dataset can be a highly effective way of quickly identifying trends and generating ideas for further analysis. 

More information on this approach can be found here: https://r4ds.had.co.nz/model-building.html 

## How have house prices in England and Wales changed since 2008? 

Data on median property prices in England and Wales is published quarterly by the Office for National Statistics, using data from the Land Registry. It is available [here]( https://www.ons.gov.uk/peoplepopulationandcommunity/housing/datasets/medianhousepricefornationalandsubnationalgeographiesquarterlyrollingyearhpssadataset09)   

The Excel document contains house prices for different administrative geographies of England and Wales, and different types of property. Here, we'll look at median prices for all property types in the c. 350 local authorities in England and Wales.  

We'll start by reading in the data and do some basic cleaning. 

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)

options(scipen = 9999)

house_prices <- read_xls("hpssadataset9medianpricepaidforadministrativegeographies.xls",
         sheet = "2a", skip = 6) %>% 
  # Converting the data from 'wide' to long
  gather(key = "date", value = "median_price", -1, -2, -3, -4) %>% 
  # Converting date to a date format using the myd() function from the lubridate package
  mutate(date = str_replace(date, "Year ending ", "") %>% myd(truncated = 1)) %>% 
  # Looking at data since 2008
  filter(date >= dmy("01-01-2008")) %>% 
  janitor::clean_names() 

# Clean data 
house_prices 

```

## Visualising changes to house prices

Let's quickly graph the changes to house prices. First, we'll look at absolute values. 
```{r, message=FALSE}
house_prices %>% 
  ggplot(aes(date, median_price)) + 
  geom_line(aes(group = local_authority_name), alpha = 0.2) + 
  geom_smooth(se = FALSE) +
  labs(x = "", y = "Median house price", title = "Median house prices in local authorities since 2008")

# Using a logarithmic scale 
house_prices %>% 
  ggplot(aes(date, median_price)) + 
  geom_line(aes(group = local_authority_name), alpha = 0.2) + 
  geom_smooth(se = FALSE) + 
  scale_y_log10() +
  labs(x = "", y = "Median house price (log scale)", 
       title = "Median House Prices in local authorities since 2008")

```

The range of house prices across England and Wales is very large, so we'll look at relative change. We'll create an index that looks at prices relative to their level in March 2008. We'll look at one visualisation for all of the locations, and then faceted by region. 

```{r, message=FALSE}
house_prices <- house_prices %>% 
  group_by(local_authority_name) %>% 
  mutate(median_price_index = (median_price/median_price[date == min(date)])*100) %>% 
  ungroup()

house_prices %>% 
  ggplot(aes(date, median_price_index)) + 
  geom_line(aes(group = local_authority_name), alpha =  0.2) + 
  geom_smooth(se = FALSE) + 
  labs(x = "", y = "Median house index (Mar 2008 = 100)", 
       title = "Median house prices in local authorities since 2008")

house_prices %>% 
  ggplot(aes(date, median_price_index)) + 
  geom_line(aes(group = local_authority_name), alpha =  0.2) + 
  geom_smooth(se = FALSE) + 
  facet_wrap(~region_country_name) +
  labs(x = "", y = "Median house index (Mar 2008 = 100)", 
       title = "Median house prices in local authorities since 2008")

```

# K-means clustering 

We'll first use k-means clustering to identify areas which have experienced similar patterns in their house price growth. I won't explain how k-means clustering works, as there are countless articles on this around the Web already. 

```{r}
# Preparing the data 
house_prices_kmeans <- house_prices %>% 
  select(local_authority_name, date, median_price_index) %>% 
  spread(date, value = median_price_index) %>% 
  as.data.frame()

hp_rownames <- pull(house_prices_kmeans, local_authority_name)

house_prices_kmeans <- house_prices_kmeans %>% select(-local_authority_name)

rownames(house_prices_kmeans) <- hp_rownames

# Identifying the best number for 'k' 
set.seed(42)

total_withiness <- map_dbl(1:10,  function(k){
  model <- kmeans(x = house_prices_kmeans, centers = k)
  model$tot.withinss
})

elbow_df <- data.frame(
  k = 1:10,
  total_withiness = total_withiness
)

ggplot(elbow_df, aes(x = k, y = total_withiness)) +
  geom_line() +
  scale_x_continuous(breaks = 1:10)

```

We'll go with 4 clusters 
```{r}

model <- kmeans(house_prices_kmeans, centers = 4)

library(broom)

model_results <- augment(model, house_prices_kmeans) %>% 
  select("local_authority_name" = 1, "cluster" = .cluster) 

house_prices <- inner_join(house_prices, model_results, by = "local_authority_name")
  
```

### Visualising the k-means results 
Visualising the house price changes per cluster shows that the k-means clustering has identified areas which have experienced different trajectories of house price growth. 

```{r}

house_prices %>% 
  ggplot(aes(date, median_price_index)) + 
  geom_line(aes(group = local_authority_name, col = cluster)) +
  modelr::geom_ref_line(h = 100) + 
  facet_wrap(~cluster)

```

Overlaying the clustering onto a map, we can see that the group with the largest increases in house prices (group 2) are located in London (with the one exception of Cambridge). The group with the second largest increases in house prices (group 4) are located around London and the South East. 

In general, the further a location is from London, the less house prices have increased since 2008. It's worth noting that locations closer to London (and those in London) also had higher house prices to start with. In these locations, prices started higher than they were elsewhere in England and Wales and have also grown faster.    

```{r, message=FALSE}
library(sf)
library(leaflet)

england_wales_map <- read_sf("Local_Authority_Districts_December_2017_Full_Clipped_Boundaries_in_Great_Britain", 
        "Local_Authority_Districts_December_2017_Full_Clipped_Boundaries_in_Great_Britain") %>% 
  select(lad17cd) %>% 
  st_simplify(dTolerance = 100) %>% 
  st_transform(crs = 4326) 

england_wales_map <- england_wales_map %>% 
  inner_join(distinct(house_prices, local_authority_code, local_authority_name, cluster), 
             by = c("lad17cd" = "local_authority_code")) 

fact_pal <- colorFactor("viridis", england_wales_map$cluster)

leaflet(england_wales_map) %>% 
  addPolygons(color = "white", weight = 1, fillColor = ~fact_pal(cluster), fillOpacity = 1, 
              popup = ~local_authority_name) %>% 
  addLegend(pal = fact_pal, values = ~cluster) 

```

# Hierarchical clustering

We'll now use hierarchical clustering to spot trends in the house prices data. While this method of clustering is different to k-means, the results should be comparable to the k-means clustering.  

```{r, warning=FALSE}
# Preparing the data 
house_prices_hclust <- house_prices %>% 
  select(local_authority_name, date, median_price_index) %>% 
  spread(date, value = median_price_index) %>% 
  as.data.frame()

rownames(house_prices_hclust) <- pull(house_prices_hclust, local_authority_name)
house_prices_hclust <- house_prices_hclust %>% select(-local_authority_name)

# Performing the clustering and joining with our house prices data 
dist_matrix <- dist(house_prices_hclust, method = "euclidean")

house_prices_hclust <- hclust(dist_matrix, method = "complete")
plot(house_prices_hclust)

```

We'll go with four clusters again. 
```{r, message=FALSE}
hclust_out <- cutree(house_prices_hclust, k = 4) %>% 
  tidy() %>% 
  select("local_authority_name" = 1, "hclust" = 2)

house_prices <- house_prices %>% inner_join(hclust_out)

```

### Visualising the hierarchical clustering results 

```{r, message=FALSE}
house_prices %>% 
  ggplot(aes(date, median_price_index)) + 
  geom_line(aes(group = local_authority_name, color = factor(hclust))) +
  geom_smooth() + 
  modelr::geom_ref_line(h = 100) +
  facet_wrap(~hclust)

```

The hierarchical clustering has identify the same trend as the k-means clustering - areas in and around London have experienced the largest increases in house prices since 2008. It's interesting that both methods of clustering identified that Cambridge has experienced a 'London-like' pattern in its house price increases since 2008.  

```{r}

england_wales_map <- england_wales_map %>% 
  inner_join(distinct(house_prices, local_authority_code, hclust), 
             by = c("lad17cd" = "local_authority_code")) 

fact_pal <- colorFactor("viridis", england_wales_map$hclust)

leaflet(england_wales_map) %>% 
  addPolygons(color = "white", weight = 1, fillColor = ~fact_pal(hclust), fillOpacity = 1, 
              popup = ~local_authority_name) %>% 
  addLegend(pal = fact_pal, values = ~hclust) 

```


